---
title: "p8105_hw3_crd2162"
author: "Caleigh Dwyer"
date: "2023-10-11"
output: github_document
---


```{r set up, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

##Problem 1: Instacart


```{r import_data}
data("instacart")
instacart = 
  instacart |> 
  as_tibble()
```

This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row representing a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart |> select(product_id) |> distinct() |> count()` products found in `r instacart |> select(order_id) |> distinct() |> count()` orders from `r instacart |> select(user_id) |> distinct() |> count()` distinct users.

Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r aisle_count}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

Next is a plot that shows the number of items ordered in each aisle. Here, aisles are ordered by ascending number of items.

```{r aisle_plot}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

Our next table shows the three most popular items in aisles `baking ingredients`, `dog food care`, and `packaged vegetables fruits`, and includes the number of times each item is ordered in your table.

```{r aisle_rank}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week. This table has been formatted in an untidy manner for human readers. Pink Lady Apples are generally purchased slightly earlier in the day than Coffee Ice Cream, with the exception of day 5.

```{r aisle_time}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  mutate(
    order_dow =
    case_match(
      order_dow,
      0 ~ "1",
      1 ~ "2",
      2 ~ "3",
      3 ~ "4",
      4 ~ "5",
      5 ~ "6",
      6 ~ "7"
    )
  ) |> 
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```


##Problem 2

In the chunk below, we will import and clean the BRFSS dataset. BRFSS is a continuous, state-based surveillance system that collects information about modifiable risk factors for chronic diseases and other leading causes of death. 

```{r clean_brfss}
data("brfss_smart2010")

brfss = 
  brfss_smart2010 |> 
  as_tibble()|> 
  janitor::clean_names() |> 
  rename(
    state = locationabbr,
    location = locationdesc) |> 
  select(year, state, location, topic, response, data_value) |> 
  filter(
    topic == "Overall Health"
    ) |> 
  mutate(
    response = 
      factor(response, levels = c("Poor", "Fair", "Good", "Very good", "Excellent"), ordered = TRUE)
  )
```

The following table shows the states with 7 or more observed locations in 2002:

```{r locations_2002}
brfss |> 
  filter (year ==2002) |> 
  group_by(state) |> 
  summarize(location_count = n_distinct(location)) |>
  filter (location_count >= 7) |>
  arrange(desc(location_count)) |>
  knitr::kable()

```

This table shows the states with 7 or more observed locations in 2010:

```{r locations_2010}
brfss |> 
  filter (year ==2010) |> 
  group_by(state) |> 
  summarize(location_count = n_distinct(location)) |>
  filter (location_count >= 7) |>
  arrange(desc(location_count)) |>
  knitr::kable()

```
 

In the following chunk, we will create a dataset limited to "Excellent" responses to the overall health question and create a new variable that calculates the mean data_value for each state by year.

```{r brfss_excellent}
brfss_excellent = 
  brfss|> 
  select(year, state, location, response, data_value) |> 
  filter(response == "Excellent")


brfss_excellent_mean =
  brfss_excellent |> 
  group_by(state, year) |> 
    mutate(
    dv_mean = mean(data_value, na.rm = TRUE)
  ) 

```

In the following chunk, we will construct a "spaghetti plot" using the brfss_excellent dataset.

```{r}
brfss_excellent_mean |> 
  ggplot(aes(x=year, y=dv_mean, group = state, color = state))+
  geom_line()

```

In the next chunk, we will produce a plot showing, for the years 2006 and 2010, the distribution of data_value for responses among locations in NY State.

```{r brfss_ny}
brfss_ny =
  brfss |> 
  filter(
    year == "2006" | year == "2010",
    state == "NY"
  ) 

brfss_ny |> 
  ggplot(aes(x=response, y=data_value, color = year))+
  geom_point()+
  facet_grid(.~year)
```


##Problem 3: Accelerometer data

In the chunk below, we import the accel and covar datasets from nhanes.

```{r import_nhanes}
accel =
  read_csv(
    "data/nhanes_accel.csv"
  ) |> 
  janitor::clean_names() |> 
  na.omit()


covar =
  read_csv(
    "data/nhanes_covar.csv",
    skip =4) |> 
  janitor::clean_names() |> 
  mutate(
    sex =
      case_match(
        sex,
        1 ~ "male",
        2 ~ "female"),
    education =
      case_match(
        education,
        1 ~ "less than high school",
        2 ~ "high school equivalent",
        3 ~ "more than high school"
      )
      ) |> 
  mutate(
    education = 
      fct_relevel(education, c("less than high school", "high school equivalent", "more than high school"))
  ) |> 
  filter(age > 21) |> 
  na.omit()
```

In the next chunk, we'll join the datasets.

```{r join_nhanes}
nhanes_combined =
left_join(x = covar, y = accel)
```

The table below shows the number of men and women in each education category:

```{r edu_sex_table}
nhanes_edu_sex =
nhanes_combined |> 
  select(education, sex) |> 
  group_by(education, sex) |> 
  summarize(n = n()) 

nhanes_edu_sex|>
  knitr::kable()
```

The chunk below produces a visualization of the table above:

```{r}
nhanes_edu_sex|> 
  ggplot(aes(x=education, y = n, fill= sex))+
  geom_bar(stat = "identity", position = "stack")+
  labs(x = "Education Level", y = "Count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We can see that the ratio of men to women at each level of education is mostly consistent (about 50/50) except that in the high school equivalent level, there is a slightly higher proportion of men than women. 

In the chunk below we will aggregate the accelerometer data:

```{r aggr_accel}
nhanes_activity =
nhanes_combined |> 
  mutate(
    total_activity = rowSums(select(nhanes_combined, min1:min999), na.rm = TRUE)
  ) |> 
  select(total_activity, age, sex, education) |> 
  view()
```

Now we will create a visualization showing the total activities against age, comparing men and women and with separate panels for each education level. 

```{r activity_plot}
nhanes_activity |> 
  ggplot(aes(x=age, y=total_activity, color = sex))+
  geom_point()+
  facet_grid(.~education)+
  geom_smooth(se= FALSE)
```

Lastly we will create a plot that shows the 24-hour activity by education level and sex.
```{r}
nhanes_minutes =
nhanes_combined |>
  pivot_longer(cols = starts_with("min"),
               names_to = "minute",
               values_to = "activity") |> 
  na.omit()
  
nhanes_minutes |> 
  ggplot(aes(x=minute, y = activity, color = sex))+
  geom_point(alpha = 0.5)+
  labs(x = "Time", y = "Activity")+
  facet_grid(.~education)+
  geom_smooth(method = lm, se=FALSE)
```

This plot shows that across education levels, there is a significant dip in activity levels in the middle of the day. This could be due to participants working sedentary jobs during the middle of the day. Activity levels are mostly consistent between males and females across time and education levels. Notably, in the "more than high school" group, male participants had a large spike in activity in the morning, whereas female participants had a large (though not quite as large) spike in activity in the afternoon. It's unclear what could be driving the gender difference in the "more than high school" group.

